Composition-climate paper

### 2015-11-12 Composition_Ecology.Rmd
Tried a new tack to get the subscripting to work.  If it doesn't it's not the end of the world.
Minor text edits in the introduction.

Worked to fix the figures in the paper (see edits in `gain_loss_plot`).

Added text to discussion.

NOTE:  The hypothesis is that we have three different types of land use/climate interactions.  This needs to be followed up in the analysis.

**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.  PDSI shows strong coherence with A_ppt_ in this region. (f) Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three tpyes of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].  Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental variable. Confounding effects may increase the variability of predictive models, but do little to the overall envelope.  Garry oak distributions in the Pacific Northwest may suffer from confounding effects.  The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental]. Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.  Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

For southern tree species, land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are clearly responding to climate change within forested landscapes, showing, for example, differences in the distributions of mature trees and seedlings [@monleon2015evidence], but modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts. For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

@@

Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.  Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C)).  The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)` since the 1895-1925 normal.  This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.

@@

loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable, loss_melt[[1]]$value))
loss_table$Gain <- loss_table$Gain / (loss_table$Loss + loss_table$Presence)
loss_table[,2:3] <- tran(loss_table[,2:3], method = 'proportion')

loss_table[,2:3] <- paste0(round(loss_table[,2:3]*100,1),'%')

@@

**Figure 3.**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Continuous presence is indicated by dark green.  For most taxa it is the central parts of their distribution that show continuous presence.  Few taxa show novel presence (purple), although taxa such as poplar, ash and basswood do show gains of over 10%.  Most taxa show significant losses in distribution (orange).*

**Table 1.** *Gain and loss of the various forest tree types of interest.*
```{r, results='as.is'}
pandoc.table(loss_table, justify = "left", style = 'simple')
```

@@

**Figure 4.** *Changes in the overall climate space in the upper midwestern United States. The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*

@@

Interstingly, since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.

Discussion
----------------

Figures
--------------

>>>>

**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].  Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient. Confounding effects may increase the variability of predictive models, but do little to the overall envelope.  Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.  The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used. Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.  Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are responding to climate change within forested landscapes, but the response is inconsistent. Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure]. Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns. For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

@@

Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.  Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C).  The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal.  This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.

@@

loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable,
                                         loss_melt[[1]]$value))
loss_table$Gain <- round(loss_table$Gain / (loss_table$Loss + loss_table$Presence)*100,1)

loss_table[,2:3] <- round(tran(loss_table[,2:3], method = 'proportion') * 100,1)

@@

**Figure 3.**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Loss is assessed where previously PLSS data indicated presence of a taxon but FIA data do not record presence (light blue). Continuous presence is indicated by dark gray, for where both FIA and PLSS cells indicate presence.  For most taxa it is the central parts of their distribution that show continuous presence.  Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*

**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era. Gain since the pre-settlement era is the number of new cells as a proportion of historical cover for the tree taxon, thus positive values are not indicative of higher overall spatial coverage. Loss and Presence columns sum to 100% and represent the proprtion of the historical range that has been lost, or shown continuous presence.*

```{r, results='as.is', echo=FALSE}
pandoc.table(loss_table, justify = "left", style = 'simple')
```

@@

**Figure 4.** *Changes in the realized climate space for genera in the upper Midwestern United States. For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.  At the same time minimum temperatures and annual precipitation have both increased.*

@@

Since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.

Discussion
----------------
This work adds to our understanding of the relationships between shifts in realized climate niches and the interacting effects of climate change and land use.

Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000]. Land use change reduces the correlational structure between species and climate [@devineau2011arid]. While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015], and in the correlational structure between species and climate. Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013niche].

From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States. However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].  In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013niche].
